---
title: "Imputation of Financial Time Series"
author: |
  | Junyan LIU and Daniel P. Palomar
  | Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: pygments  
    toc: yes
    toc_depth: 2
  prettydoc::html_pretty:
    theme: tactile
    highlight: vignette
    toc: yes
    toc_depth: 2
csl: ieee.csl
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Imputation of Financial Time Series}
  %\VignetteKeyword{imputation, time series, AR, random walk}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knit_hooks$set(pngquant = hook_pngquant)
# Help on bookdown: https://bookdown.org/yihui/bookdown/
# rmarkdown::render("ImputeFinancialTimeSeries.Rmd", "prettydoc::html_pretty")
```

-----------
> This vignette illustrates the usage of the package [`imputeFin`](https://CRAN.R-project.org/package=imputeFin) 
for imputation of missing values in financial time series that fit a random walk model or an autoregressive (AR)
model.


# Installation
The package can be installed from [CRAN](https://CRAN.R-project.org/package=imputeFin) or [GitHub](https://github.com/dppalomar/imputeFin):
```{r, eval=FALSE}
# install stable version from CRAN
install.packages("imputeFin")

# install development version from GitHub
devtools::install_github("dppalomar/imputeFin")

# Getting help
library(imputeFin)
help(package = "imputeFin")
?imputeAR1Gaussian
```



# Usage of the package
## Estimation and imputation for incomplete AR(1) time series
We start by loading the package:
```{r, message = FALSE}
library(imputeFin)
```

We download the adjusted price of S&P 500 index from 2012-01-01 to 2015-07-08, and compute the log prices. Some log prices are deleted to generate missing values.

```{r, message = FALSE}
library(xts)
library(quantmod)
y_orig <- log(Ad(getSymbols("^GSPC", from = "2012-01-01", to = "2015-07-08", auto.assign = FALSE)))
n <- nrow(y_orig)
miss_pct <- 0.2 # the percentage of missing values
n_miss <- floor(miss_pct*n)
index_miss <- round(n/2) + 1:n_miss
y <- y_orig
y[index_miss] <- NA
```

Then we use the function `estimateAR1t()` to estimate the parameters of Student's t AR(1) model from this incomplete time series.

```{r, message = FALSE}
estimation_result <- estimateAR1t(y, random_walk = FALSE)
estimation_result$phi0
estimation_result$phi1
estimation_result$sigma2
estimation_result$nu

imputation_result <- imputeAR1t(y, n_sample = 3, random_walk = FALSE)
```

Next, we use the function `imputeAR1t()` to impute the missing values in the incomplete time series.
 
```{r}
# we generate n_sample = 2 imputed time series
imputation_result <- imputeAR1t(y, n_sample = 2, random_walk = FALSE)
y_imputed <- imputation_result$y_imputed

```

The functions `estimateAR1t()` and `imputeAR1t()` can also deal with multivariate incomplete time series. They will do estimatation and imputation for each univariate time series seperately.

```{r}
stocks <- lapply(c("^GSPC","^HSI"), function(sym) {
  log(Ad(getSymbols(sym, from = "2018-01-01", to = "2019-01-01", auto.assign = FALSE)))
})
Y_orig <- do.call(merge, stocks)
n <- nrow(Y_orig)
miss_pct1 <- 0.1 # the percentage of missing values
miss_pct2 <- 0.2
index_miss1 <- sort(sample(2:(n - 1), n * miss_pct1))
index_miss2 <- sort(sample(2:(n - 1), n * miss_pct2))
Y <- Y_orig
Y[index_miss1, 1] <- NA
Y[index_miss2, 2] <- NA

estimation_result <- estimateAR1t(Y, random_walk = FALSE)
imputation_result <- imputeAR1t(Y, n_sample = 2, random_walk = FALSE)
```

The functions `estimateAR1Gaussian()` and `imputeAR1Gaussian()` can be used in a similar way. Instead of the Student's t AR(1) model, they fit the Gaussian AR(1) model to time series.

```{r, message = FALSE}
estimation_result_g <- estimateAR1Gaussian(y,  random_walk = FALSE)
estimation_result_g$phi0
estimation_result_g$phi1
estimation_result_g$sigma2

imputation_result_g <- imputeAR1Gaussian(y, n_sample = 2,  random_walk = FALSE)
```


# Comparison with other packages
We compare with the existing package `imputeTS`. We plot the imputed time series obtained by  functions in package `imputeTS` and our function `imputeAR1t()`，respectively.

```{r}
# impute using package imputeTS 
library("imputeTS")
y_imputed_km = na.kalman(y)
y_imputed_random = na.random(y)
y_imputed_linear = na.interpolation(y, "spline")
index_miss_bis = index_miss
par(mfrow=c(2,2))
#plot 1
{ plot(y, main="orignal")
  lines(y_orig[index_miss_bis], col="blue", lwd=2) }
#plot 2
{ plot(y, main="na.kalman")
  lines(y_imputed_km[index_miss_bis], col="blue", lwd=2) }
#plot 3
{ plot(y, main="na.random")
  lines(y_imputed_random [index_miss_bis], col="blue", lwd=2) }
#plot 4
{ plot(y, main="na.interpolation")
  lines(y_imputed_linear[index_miss_bis], col="blue", lwd=2) }

# par(mfrow=c(2,2))
# #plot 1
# { plot(y, main="orignal")
#   lines(y_orig[index_miss_bis], col="blue", lwd=2) }
# #plot 2
# { plot(y, main="imputeAR1t")
#   lines(y_imputed[index_miss_bis, 1], col="blue", lwd=2) }
# #plot 3
# { plot(y, main="imputeAR1t")
#   lines(y_imputed[index_miss_bis, 1], col="blue", lwd=2) }
# #plot 4
# { plot(y, main="imputeAR1t")
#   lines(y_imputed[index_miss_bis, 1], col="blue", lwd=2) }
```